---
title: "ITS_report"
author: "Elena D. Díaz Santiago"
date: "2024-11-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      comment = "",
                      eval=  TRUE,
                      include = TRUE)
```

```{r }
library(phyloseq)
library(vegan)
library(pairwiseAdonis)
library(microbiome)
library(microbiomeMarker)
library(qiime2R)
library(ggplot2)
library(ggpubr)
library(ggVennDiagram)
library(gridExtra) 
library(kableExtra)
library(dplyr)
library(magrittr)
library(metagenomeSeq)
library(microeco)
library(file2meco)
library(multcompView)
```

```{r echo=FALSE}
#load own functions
source("C:/Users/elenad/Documents/proyectos/scripts/functions.R")
```


# Experimental design

The objective of this study is to analyze whether soil microbial communities influence the germination of *Quercus ilex*. To this end,we have compared the structure of the communities of three locations in Spain with different climate conditions: Somiedo (Asturias), Ronda (Malaga), and Abla (Almeria)

The hypothesis is that the microbial communities living in soils beneath the trees could be positively affecting germination. 
It is proposed to perform an analysis of the diversity and composition of microbial communities in each soil type using techniques such as amplicon sequencing (e.g., using markers like the ITS rRNA gene for Fungi).
  
#Load Data  
Sequences were analyse with the software QIIME2. Now we import the generated dataset to perform downstream analysis.  
  
```{r }
physeq <-qza_to_phyloseq(features="taxonomy/len1/final_FeatureTable.qza", 
                         tree="phylogeny/len1/rooted-tree.qza", 
                         taxonomy="taxonomy/len1/taxonomy.qza", 
                         metadata="metadata.txt")
```


# Positive control - Mock Analysis 

As a quality control measure, a positive control with known microorganisms was introduced. Specifically, it contains two fungi:
- ***Saccharomyces cerevisiae***  
- ***Cryptococcus neoformans***


Therefore, the first thing we are going to check is that we have been able to taxonomically identify these microorganisms in the mock samples.

```{r mocktaxonomy}
physeq@sam_data$SampleID <- row.names(physeq@sam_data)
#obtain mock table
mock_ps <- subset_samples(physeq, SampleID=="MockZymo106")
#remove otus == 0
mock_ps.clean = prune_taxa(taxa_sums(mock_ps) > 1, mock_ps)
#agglomerate taxonomic results to Genus level
mock_tg <- tax_glom(mock_ps.clean, taxrank="Genus")
mock_tf <- tax_glom(mock_ps.clean, taxrank="Family")

#obtain taxonomic table
kable(mock_tg@tax_table[,1:6], row.names = FALSE, caption = "Mock taxonomy at Genus level.")
```

# Filter data

```{r FilterData}
#remove mock sample
physeq_wo_mock <- subset_samples(physeq, !SampleID %in% c("MockZymo106"))
#remove ASVs == 1
physeq.clean <- prune_taxa(taxa_sums(physeq_wo_mock) > 1, physeq_wo_mock)

#Remove otus not annotated Phylum level
physeq.clean <- subset_taxa(physeq.clean, Phylum != "NA")

#Clean genus list
genus_l <- get_taxa_unique(physeq.clean, taxonomic.rank="Genus")
remove <- c(NA, "uncultured")
genus_l <- genus_l[! genus_l %in% remove]
```

# Obtain the number of Genus

```{r ObtainGenusData}
#agglomerate taxonomic results to Genus level
ps.genus_all <- tax_glom(physeq.clean, taxrank="Genus")
#Obtain all the genera
genus_l <- get_taxa_unique(ps.genus_all, taxonomic.rank="Genus")
#remove unasigned genera
genus_to_remove <- c(NA, "uncultured")
genus_l <- genus_l[!genus_l %in% genus_to_remove]
ps.genus <- subset_taxa(ps.genus_all, Genus %in% genus_l)

genus_df <- psmelt(ps.genus)

#Remove abundance 0
genus_df <- genus_df[!genus_df$Abundance == 0,]
#Count the number of Genera per sample
genus_count_sample_df <- genus_df %>%
  group_by(SampleID) %>%
  summarise(Genus_nb = length(unique(Genus)))
genus_count_sample_df$Site <- physeq.clean@sam_data$Site

#ANOVA
anova <- aov(Genus_nb ~ Site, data = genus_count_sample_df)
summary(anova)

#tukey multiple comparison test (ANOVA post hoc)
tukey <- TukeyHSD(anova)

letters <- multcompLetters4(anova, tukey)

letters_df <- data.frame(Site = names(letters$Site$Letters), Letter = toupper(as.character(letters$Site$Letters)))

#Mean and SD Genera number by site
genus_count_df_fungi <- genus_count_sample_df%>%
  group_by(Site) %>%
  summarise(Genus_mean = mean(Genus_nb), SD = sd(Genus_nb))

kable(genus_count_df_fungi, row.names = FALSE, caption = "Number of genera identified per experimental condition. The table presents the total number of genera detected in each experimental Site, without considering their abundances. This provides an overview of the taxonomic diversity present in each condition based solely on the presence of genera.")
```

```{r genusbarplot}
genus_count_df_fungi <- merge(genus_count_df_fungi, letters_df, by = "Site")
genus_count_df_fungi$Type <- rep("Fungi", nrow(genus_count_df_fungi))

ggplot(genus_count_df_fungi, aes(x=Site, y=Genus_mean)) + 
  geom_bar(stat="identity", position=position_dodge()) +
  geom_errorbar(aes(ymin=Genus_mean, ymax=Genus_mean+SD), width=.2, position=position_dodge(.9)) +
  geom_text(mapping = aes(label = Letter, y = Genus_mean * 1.3, size=12), 
              position = position_dodge(width = 0.9), show.legend = F, fontface = "bold") +
  theme_classic()
```

# Obtain the number of Phylum

```{r ObtainPhylumData}
#agglomerate taxonomic results to Phylum level
ps.phylum <- tax_glom(physeq.clean, taxrank="Phylum")
phylum_df <- psmelt(ps.phylum)

#Remove abundance 0
phylum_df <- phylum_df[!phylum_df$Abundance == 0,]
#Count the number of Genera per sample
phylum_count_sample_df <- phylum_df %>%
  group_by(SampleID) %>%
  summarise(Phylum_nb = length(unique(Phylum)))
phylum_count_sample_df$Site <- physeq.clean@sam_data$Site

#ANOVA
anova <- aov(Phylum_nb ~ Site, data = phylum_count_sample_df)
summary(anova)

#tukey multiple comparison test (ANOVA post hoc)
tukey <- TukeyHSD(anova)

#Mean and SD Genera number by site
phylum_count_df <- phylum_count_sample_df%>%
  group_by(Site) %>%
  summarise(Phylum_mean = mean(Phylum_nb), SD = sd(Phylum_nb))

kable(phylum_count_df, row.names = FALSE, caption = "Number of Phylum identified per experimental condition. The table presents the total number of Phylum detected in each experimental Site, without considering their abundances. This provides an overview of the taxonomic diversity present in each condition based solely on the presence of Phylum.")
```

# Alpha diversity
```{r AlphaDiversity}
metrics <- c("Shannon", "Simpson", "Observed")
comparisons <-list(c("Ronda", "Somiedo"))
#STABLISH THE STATISTIVAL METRICS TO CALCULATE IF THE DIVERSITY IS SIFNIFICANT BETWEEN GROUPS
stat_metrics <- c("wilcox.test")

#OBTAIN ALPHA DIVERSITY PLOT
alpha_fungi <- alpha_diversity_plot(physeq.clean, "Site", metrics, comparisons, stat_metrics, asterisk = TRUE, " ")
alpha_fungi

#alpha diversity table
alpha.tb <- estimate_richness(physeq.clean, measures=metrics)
alpha.tb$Site <- physeq.clean@sam_data$Site
write.table(alpha.tb, "fungi_alpha_div_tb.tsv", sep = "\t", quote = FALSE)
kable(alpha.tb)
```

# Normalization - metagenomeSeq  

Generally, sequencing data sets are not of the same size. So, to make relatively fair comparisons between sets, we need to normalize the counts. It is necessary to prepare the data in a way that improves the comparability of the samples (e.g. alpha diversity, beta-diversity). There are different methods to perform normalization, but in this case we are using Cumultive Sum Scaling with metagenomeSeq package.

```{r PhyloseqNormalizarion}
#cumulative sum scaling
physeq.css <- physeq.clean

ps_metagenomeseq <-phyloseq_to_metagenomeSeq(physeq.clean)
obj_trans<-cumNormMat(ps_metagenomeseq)

otu_table(physeq.css) <- otu_table(obj_trans, taxa_are_rows = TRUE)
```


## Beta Diversity

```{r BraycurtisDistanceCalculation}
distance <- "bray"

set.seed(1)

bray_distance <- phyloseq::distance(physeq.css, method = distance)
meta <- data.frame(sample_data(physeq.css))
```


```{r BrayPermanova}
bray_permanova <- adonis2(bray_distance ~ Site, data = meta, permutations = 999, method = distance)

bp_pvalue <- paste("P-value", bray_permanova$`Pr(>F)`[1])
bray_permanova_text <- paste("PERMANOVA", bp_pvalue, sep = "\n")

bray_permanova
```


***PCoA - Bray Curtis*** 

```{r BraycurtisPCoA, fig.cap="Beta diversity analysis using Bray-Curtis dissimilarity visualized by PCoA. The plot represents the principal coordinates analysis (PCoA) based on Bray-Curtis distances, illustrating the dissimilarity between samples across experimental groups. A PERMANOVA test was conducted to assess statistical differences, with the p-value displayed on the graph."}
method <- "PCoA"

pcoa_bray <- betadiversity_analysis(physeq.css, method, distance)
beta_fungi <- plot_distance(physeq.css, pcoa_bray, "Site", elipse = TRUE) + annotate("text" ,x=-0.05,y=0.43, label = bray_permanova_text, size = 5) 
beta_fungi
```


# Relative Abundance

```{r calculateRelativeAbundance}
#filter NA and unclutured genus
ps_genus <- subset_taxa(physeq.clean, Genus %in% genus_l)

#Convert to relative abundance according to a taxonomic level -> Genus
genus_rel_abun <- relative_abundance_by_taxa(ps_genus, "Genus")
#physeq object
genus_rel_ps <- genus_rel_abun[[1]]
#data.frame
genus_rel_df <- genus_rel_abun[[2]]
names(genus_rel_df)[3] <- paste("Relative_Abundance")

```

## Top 20 by Sample

```{r Top20, fig.width=10, fig.height=6}
#Filter top 20
top_taxa <- top_taxa(genus_rel_ps, n = 20)
filter_top_ps <- prune_taxa(top_taxa, genus_rel_ps)
genus_top_filter.df <- phyloseq::psmelt(filter_top_ps)
names(genus_top_filter.df)[3] <- paste("Rel_Abundance")
top_sample_fungi <- plot_abundance_by_sample(genus_top_filter.df, "SampleID", "Rel_Abundance", "Genus", "SampleID") + facet_wrap(~Site, scales = "free") + theme(strip.text = element_text(size = 14))
top_sample_fungi
```

## > 1% Abundance

```{r 1RelAbundance, fig.width=10, fig.height=6}
ps_genus_1 <- filter_taxa(genus_rel_ps, function(x) mean(x) > 1,TRUE)
genus_1_filter.df <- phyloseq::psmelt(ps_genus_1)
names(genus_1_filter.df)[3] <- paste("Rel_Abundance")
plot_abundance_by_sample(genus_1_filter.df, "SampleID", "Rel_Abundance", "Genus", "SampleID") + facet_wrap(~Site, scales = "free") + theme(strip.text = element_text(size = 14))

```


## Top 20 By Genera
  
**Top 20**  
```{r Top20Genera, fig.width=8}

top_genus_rel_sumarize <- genus_top_filter.df %>%  group_by(Site, Genus) %>% 
  summarise(Relative_Abundance=mean(Rel_Abundance, na.rm=TRUE), sd=sd(Rel_Abundance, na.rm=TRUE))

top_fungi <- plot_abundance_by_taxa(top_genus_rel_sumarize, "Genus", "Relative_Abundance", "Site")
top_fungi

```
  
**Top 20 by Site**  

```{r Top20BySite, fig.width=8}
#Top 20 by site
ps_rel_ronda <- subset_samples(genus_rel_ps, Site %in% "Ronda")
top_taxa_ronda <- top_taxa(ps_rel_ronda, n = 20)
ps_top_taxa_ronda <- prune_taxa(top_taxa_ronda, ps_rel_ronda)
ronda_top_taxa.df <- phyloseq::psmelt(ps_top_taxa_ronda)

ps_rel_somiedo <- subset_samples(genus_rel_ps, Site %in% "Somiedo")
top_taxa_somiedo <- top_taxa(ps_rel_somiedo, n = 20)
ps_top_taxa_somiedo <- prune_taxa(top_taxa_somiedo, ps_rel_somiedo)
somiedo_top_taxa.df <- phyloseq::psmelt(ps_top_taxa_somiedo)

top_taxa.df <- rbind(ronda_top_taxa.df, somiedo_top_taxa.df)
names(top_taxa.df)[3] <- paste("Rel_Abundance")


top_20_genus_rel_sumarize <- top_taxa.df %>%  group_by(Site, Genus) %>% 
  summarise(Relative_Abundance=mean(Rel_Abundance, na.rm=TRUE), sd=sd(Rel_Abundance, na.rm=TRUE))

topsite_fungi <- plot_abundance_by_taxa(top_20_genus_rel_sumarize, "Genus", "Relative_Abundance", "Site")
topsite_fungi
```


# Functional Traits Analysis


This analysis is carried out thanks to the R microeco package, which searches for the functions related to prokaryotic communities using the FAPROTAX database. Subsequently, it performs a differential abundance analysis of the routes found depending on the conditions that are established.

```{r DAA-FTA, fig.width=8, fig.height=7}
#Create microeco object with Ronda and Somiedo Samples
meco_fung <- phyloseq2meco(genus_rel_ps)
t1_fung <- trans_func$new(meco_fung)

#Search in FUNGuild
t1_fung$cal_spe_func(fungi_database = "FUNGuild", FUNGuild_confidence = c("Highly Probable", "Probable", "Possible"))
#obtain abundance
t1_fung$cal_spe_func_perc(abundance_weighted = TRUE)
#obtain percentage
t1_fung$trans_spe_func_perc()

# clone meco_fung for differential test
tmp_fung <- clone(meco_fung)

# we need assign the table to taxa_abund manually
tmp_fung$taxa_abund$func <- as.data.frame(t(t1_fung$res_spe_func_perc), check.names = FALSE)
# use trans_diff class to perform differential test
t2_fung <- trans_diff$new(dataset = tmp_fung, method = "wilcox", group = "Site", taxa_level = "func", alpha = 0.05)

sig_pathways <- t2_fung$res_diff[t2_fung$res_diff$P.adj < 0.05,]
pathway_l <- unique(sig_pathways$Taxa)
abund_t_fung <- t2_fung$res_abund[t2_fung$res_abund$Taxa %in% pathway_l, ]
abund_t_fung$Taxa <- as.character(abund_t_fung$Taxa) 

textdf_fung <- obtain_significance_df(sig_pathways, abund_t_fung)
fta_bpl_fung <- fta_abundance_barplot(abund_t_fung, textdf_fung, "Site")
fta_bpl_fung
```